---
title: "Reproducible Research Project1"
author: "Mary T."
date: "April 9, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## (1) Loading and preprocessing the data
## download file
```{r}
library(plyr)
library(ggplot2)
file_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
file_name <- "repdata_data_activity.zip"
download.file(file_url, file_name, method = "curl")
unzip(file_name)

```
## Read file
```{r}
activity <- read.csv('./activity.csv', header=TRUE, na.strings="NA")
str(activity)
```

## Format date variable to date class and interval variable to factor class
```{r}
activity$date <- as.Date(activity$date)
activity$interval <- as.factor(activity$interval)
```
## (2) What is mean total number of steps taken per day?
## Plot a histogram of total steps taken each day

# Total steps taken each day
```{r}
activity_1 <- ddply(activity, c("date"), summarize,
                     steps  = sum(steps,na.rm = TRUE)
 )
```
# Mean and median of the total number of steps taken each day
```{r}
steps_mean = mean(activity_1$steps)
steps_median = median(activity_1$steps)

sprintf("Mean number of steps: %s", steps_mean)
sprintf("Median number of steps: %s", steps_median)
```
```{r}
ggplot(activity_1, aes(x=steps)) + 
    geom_histogram(binwidth = 2000, color="blue", fill="white")+
    xlab("Total steps per day") +     ylab("Frequency") +
    ggtitle("Total Number of Steps Taken Each Day") +
    geom_vline(aes(xintercept = steps_mean, color="mean"), size = 0.7) +
    geom_vline(aes(xintercept = steps_median, color="median"), size = 0.7)
```


## (3) What is the average daily activity pattern?
# Average steps taken each day
```{r}
activity_2 <- ddply(activity, c("interval"), summarize,
                     average_steps  = mean(steps,na.rm = TRUE)
 )
```
# Make a time-series plot of the 5-minute interval (x-axis) and the average number of steps taken (y-axis)
```{r}
x <- activity_2$interval
y <- activity_2$average_steps
plot(x,y, type="l", xlab = "5 Minute Intervals", ylab = "Average Number of Steps Taken", main = "The Average Daily Activity Pattern", col="blue")
```
# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_steps <- activity_2$interval[which.max(activity_2$average_steps)]
sprintf("The 5-minute interval with the maximum number of steps on average is: %s", max_steps)
```
## (4) Impute missing values
# Calculate and report the total number of missing values (NA)
```{r}
na_values <- is.na(activity[,1])
sprintf("Number of missing values: %s", sum(na_values))
```
# Devise a strategy for filling in all the missing values in the dataset, e.g. use the mean, median for the day ,or the mean for the interval
# Using the mean steps for the 5-minute intervals, create new dataframe 
```{r}
activity_3 <- merge(activity, activity_2, by = "interval")
```
#Create a new dataset that is equal to the original dataset but with the missing values filled in with mean steps
```{r}
activity_3[na_values, "steps"] <- activity_3[na_values, "average_steps"]
new_dataset <- activity_3[,3:2]

activity_4 <- ddply(new_dataset, c("date"), summarize,
                     steps  = sum(steps,na.rm = TRUE)
 )
```
#Make a histogram of the total steps take each day. Calculate the mean and median. Do the values differ from the first part of the assignment?
```{r}
ggplot(activity_4, aes(x=steps)) + 
    geom_histogram(binwidth = 2000, color="black", fill="white")+
    xlab("Total steps per day") +     
    ylab("Frequency") +
    ggtitle("Total Number of Steps Taken Each Day") +
    geom_vline(aes(xintercept = mean(activity_4$steps), color="mean"), size = 1.0) +
    geom_vline(aes(xintercept = median(activity_4$steps), color="median"), size = 1.0)
```
```{r}
sprintf("New mean number of steps: %s", mean(activity_4$steps))
sprintf("New median number of steps: %s", median(activity_4$steps))
```
##Answer: Mean and median steps are similar to the original values. The shape of the histogram is also similar to the original.


## (4) Are there differences in activity patterns between weekends and weekdays?
# Create a new factor variable in the dataset with two variables - weekday and weekend
```{r}
activity_3$weekdays <- weekdays(activity_3$date)

activity_5 <- ddply(activity_3, c("interval","weekdays"), summarize,
                     average_steps  = mean(steps,na.rm = TRUE)
 )
```
# Make a panel plot containing a time-series plot of the 5-minute interval (x-axis) indicating whether a given date is a weekday or a weekend (y-axis)
```{r}
x <- activity_5$weekdays
y <- activity_5$average_steps

qplot(interval, average_steps, data = activity_5, 
	facets = weekdays~., 
	xlab = "5 Minute Intervals", 
	ylab = "Average Number of Steps Taken", 
	main = "The Average Daily Activity Pattern") +
	geom_smooth()
```





